---
title: "BIO302 exam Part A"
author: "Liem Oeastraadt-Jennings"
format: html
editor: visual
date: 29.06.2023
---

## Part A - reproducibility

1.  Continue with your attempt to reproduce part of a paper from archived data from the first practical. Archive the code and text on GitHub and include the link to your repo in your answer here.

```{r}
library(ggplot2) 
library(tidyverse) 
library(readr) 
library(here) 
library(usethis) 
use_git_config(   user.name = "liemdominic",    
                  user.email = "lje006@uib.no"   ) 
usethis::create_github_token() 
gitcreds::gitcreds_set()
usethis::use_git()
git_vaccinate()
use_github()
```

```{r}
## Importing data coral_cover.csv
coral_cover_data <- read.csv(here("coral_cover.csv")) |>
janitor::clean_names()

view(coral_cover_data)

# Plotting
coral_cover_plot <- ggplot(coral_cover_data, #give the plot a name
       aes(x = site, 
           y = mean_coral_cover, 
           fill = as.factor(year))) + #continous scale issue when year not a factor
  geom_col(position = "dodge") + #to get col's side by side
  labs(x = "Site", 
       y = "Coral Coverage (%)", 
       fill = "Year") +
  scale_fill_manual(values = c("blue", "red")) + #figure in paper had blue and red colours, not exactly sure which shade used.
  theme_bw() +
  scale_x_continuous(breaks = unique(as.numeric(coral_cover_data$site))) #to get a continous scale on x axis

coral_cover_plot
```

I tried to reproduce figure 2c) from the paper Cook et al., 2022. Looking at the data they have published which i downloaded from: https://datadryad.org/stash/dataset/doi:10.5061/dryad.9ghx3ffk7 - I can see that they have cleaned the data a lot, and that only the specific data for the figure is published. The data for the figure is called "coral_cover.csv". I am not sure if the data has been cleaned with code or if it has been cleaned using another data managing program. To try to understand how they had cleaned the data i read the methods (Cook et al., 2022) and readme.txt file, also available for download. According to the methods and readme.txt, all data has been analysed with R, but the data cleaning procedure is not specifically mentioned.

The colour and legend are not the same as in the article figure, but those are reproducible with known choice of colour and legend location coordinates. I did not use unneccessary time trying to reproduce these.

Reproducing the figure went well, but understanding how the data cleaning process was done was not possible (Cook et al., 2022). Understanding the whole figure by itself is not possible, other than understanding that there is a visual difference between mean coral coverage at the times surveyed. Visual surveys by SCUBA diving was the method used and mean percentage coverage across the site-specific quadrants was calculated using scleractinian coral quadrat data. To reproduce the figure completely i would need the coordinates for the offset of legend and the colour scheme they chose. In the data coral_cover.csv, data points are from year 1976 and 2018, but in the figure 2c (Cook et al., 2022), the year's in the legend says: 1975 and 2018. Figure text explains that the visual survey was done in the years 1975-1976 and 2018. Why the legend shows only year 1975 i do not know, as the data shows only year 1976 and 2018. An explenation of this would leave less uncertainty to what this means.

16 data points for 1976 and 16 data points for 2018 were available in the coral_cover.csv file. Data was represented in year, site and mean coral coverage (%). I used janitor::clean_names to format the data column names. I used "here" to create a pathway that does not need Rstudio and creates a file path that locates the file from any other project root. I have commented the steps in my code so that my code is hopefully understandable by others.

Not sure how to cite the data.

Sources:

Cook, K.M. et al. (2022) "A community and functional comparison of coral and reef fish assemblages between four decades of coastal urbanisation and thermal stress," Ecology and evolution, 12(3), pp. e8736-n/a.
